---
title: "TWAS gene Gene Set Analysis"
author: "Shuyang Yao (shuyang.yao@ki.se)"
date: "2023-11-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
  
- Purpose: Here we perform the gene set analysis (GSA) for the TWAS genes at TWAS-FDR≤0.3 level.  
- Hypergeometric test was used to do the GSA.  
- Output:  
 = Figure 4 panels K, L, and M (treemap plot for GO:BP in Excitatory, Inhibitory, and NonNeuronal cell types), at GSA test FDR≤0.05  
 = Figure S18 (dot plot for GO:BP, CC, and MF for each cell type), at GSA test FDR≤0.15
  
--- script starts ---  
  
GO Pathway enrichment using hypergeometric test:  
```{r}
library(data.table)
library(tidyverse)
library(ggplot2)
library(rstatix)

# hypergeometic test

#====== gene set analysis in this context ======;
  # N = Ngene = genes in geneset_background (ie brain expressed)
  # R = VLIST = genes implicated in TWAS;
  # n = nVar  = genes in genesets
  # x = xVar  = overlap of R and n
  # PH= phyper(x-1, R, N-R, n, lower.tail=FALSE)

#====== background geneset
# geneMatrix, limit to unique gene_name, protein-coding, autosomal
# have to look at 1:1 mouse-human as all pathways are human
a <- fread("./workflow/geneMatrix.tsv.gz") %>%
  filter(dup_gene_name == FALSE & gene_type == "protein_coding") %>%
  filter(hg19g0 != "chrX" & hg19g0 != "chrY" & hg19g0 != "chrM") 

#=== background genes; brain expressed genes
background <- a %>% 
  filter(is.na(BrainCortex)==F & BrainCortex>0) %>%
  distinct(ensgid) 
Ngene <- as.numeric(nrow(background))

#=== get genesets, INNER join to limit to background
b <- fread("./workflow/genesetsB.tsv.gz", stringsAsFactors=FALSE, data.table=FALSE) # gene list of GO pathways
genesets0 <- inner_join(background, b, by = "ensgid")
genesets <- genesets0 %>%
  filter(group %in% c("gene-ontology","synaptic-gene-ontology","GWAS"))
rm(a,b)

#############################
#############################
# TWAS per cell type
TWAS.FDRthresh=0.3
df.cluster <-fread("T13_Table_TWAS_full_results.tsv") %>% # supplementary table 13
  filter(level=="16 cell types") %>%
  filter(FDR_fr_Padj<=TWAS.FDRthresh) %>%
  rename(lvl=level, CT=cell_type) %>%
  arrange(lvl,CT)
CT <- unique(df.cluster$CT)
#############################
#############################

# enrichment analysis:
GSA.FDRthresh=0.05
for(i in 1:length(CT)){
  mydatname <- CT[i]
  tmp <- df.cluster %>% distinct(CT,lvl) %>% filter(CT==mydatname)
  mylvl <- as.character(tmp[1,"lvl"])
  b <- df.cluster %>% 
    filter(CT==mydatname) %>% 
    select(ensgid) %>%
    mutate(mydat=1)
  
  #=== INNER join with test list background, count "R"
  c <- inner_join(background, b, by = "ensgid") 
  R <- as.numeric(sum(c$mydat))
  
  #=== get the overlap, LEFT to preserve full geneset
  # add count vars, fix NAs
  j <- left_join(genesets, c, by = "ensgid") %>%
    select(-ensgid) %>%
    mutate(cnt = 1) %>%
    replace_na(list(mydat=0))
  
  #=== get P-hyper
  l <- j %>%
    group_by(group, subgroup, genesetID) %>% 
    summarise(nGenes=sum(cnt),
              Ngene = Ngene,
              R = R,
              x=sum(mydat)) %>%
    ungroup() %>%
    filter(nGenes>=0)
  
  l$PH <- phyper(l$x-1, l$R, l$Ngene-l$R, l$nGenes, lower.tail=FALSE)
  
  l <- l %>% 
    group_by(group) %>%
    adjust_pvalue(p.col = "PH", output.col = "P.fdr.group", method = "fdr") %>%
    ungroup()
  
  #=== select sig. results
  g <- l %>% filter(P.fdr.group<=GSA.FDRthresh) 
  g$annotation <- mydatname
  g$annotation_lvl <- mylvl

  #=== store results in one data frame (differentiated by "annotation" column)
  if(i==1){
    myres <- g
  }else{
    myres <- rbind(myres, g)
  }
}

#- get meanings of genesets
genesetsA <- fread("./workflow/genesetsA.tsv.gz")

myres <- myres %>%
  left_join(genesetsA, by="genesetID") %>% 
  filter(subgroup!="HPO")  %>%
  separate(col=geneset, sep=" ", into=c("ID","description")) %>%
  mutate(annotation_broad15CT=case_when(substr(annotation,1,3)=="Exc" ~ "Excitatory", # combine 16 cell types into 3 classes
                                    substr(annotation,1,3)=="Inh" ~ "Inhibitory",
                                    substr(annotation,1,3)=="Oli" ~ "Olig_OPC")) %>%
  select(annotation, annotation_lvl, annotation_broad15CT, subgroup, ID, description, P.fdr.group, nGenes, Ngene, R, x, PH)

#- output
outfile <- "./10_TWAS/output/fig4KLM/df_15CT_TWAS0.3_GSA0.05_GO_group3class.tsv"
fwrite(myres, file=outfile, sep="\t", col.names = T)

```
  
rrvgo plot, Fig 4KLM: TWAS FDR≤0.3, GSA FDR≤0.05  
```{r}
library(data.table)
library(tidyverse)
library(ggplot2)
library(rrvgo)

# set path
OUTDIR <- "./10_TWAS/output/fig4KLM/"
# read data
myres <- fread(paste(OUTDIR,"df_15CT_TWAS0.3_GSA0.05_GO_group3class.tsv",sep=""))

# plot rrvgo
for (myCT in c("Excitatory","Inhibitory","NonNeuronal")) {
  dat <- myres %>% filter(annotation_broad15CT==myCT)
  scores <- setNames(-log10(dat$P.fdr.group), dat$ID)
  
  #- get similarity matrix, for BP, CC, and MF separately
  simMatrix.BP <- calculateSimMatrix(dat$ID,
                                           orgdb="org.Hs.eg.db",
                                           ont=c("BP"),
                                           method="Rel")
  # input of plots
  reducedTerms.BP <- reduceSimMatrix(simMatrix.BP,
                                           scores,
                                           threshold=0.7,
                                           orgdb="org.Hs.eg.db")
  
  # plot
  plotname <- paste(OUTDIR,"fig4_",myCT,"_BP.pdf", sep="")
  pdf(file=plotname)
  treemapPlot(reducedTerms.BP)
  dev.off()
}


```
  
dop plot, FigS18: TWAS FDR≤0.3, GSA FDR≤0.15  
```{r}
library(data.table)
library(tidyverse)
library(ggplot2)
library(readxl)
library(rrvgo)

MYOUTDIR="./10_TWAS/output/figS18/"

#- specify threshold
TWAS.FDRthresh=0.3
GSA.FDRthresh=0.15
lvl="15CT"

df.color <- read_excel("./workflow/Cell_type_colors.xlsx")%>%
  distinct(`Cell type (15)`, `Cell type (15) color`)
colnames(df.color) <- c("CellType","CT_color")
df.color$CellType <- gsub(" ", "_", df.color$CellType)

#- run rrvgo
a <- fread("./workflow/geneMatrix.tsv.gz") %>%
  filter(dup_gene_name == FALSE & gene_type == "protein_coding") %>%
  filter(hg19g0 != "chrX" & hg19g0 != "chrY" & hg19g0 != "chrM") 

#=== background genes; brain expressed genes
background <- a %>% 
  filter(is.na(BrainCortex)==F & BrainCortex>0) %>%
  distinct(ensgid) 
Ngene <- as.numeric(nrow(background))

#=== get genesets, INNER join to limit to background
b <- fread("./workflow/genesetsB.tsv.gz", stringsAsFactors=FALSE, data.table=FALSE) # gene list of GO pathways
genesets0 <- inner_join(background, b, by = "ensgid")
genesets <- genesets0 %>%
  filter(group %in% c("gene-ontology","synaptic-gene-ontology","GWAS"))
rm(a,b)

#=== TWAS per cell type
df.cluster <-fread("T13_Table_TWAS_full_results.tsv") %>% # supplementary table 13
  filter(level=="16 cell types") %>%
  filter(FDR_fr_Padj<=TWAS.FDRthresh) %>%
  rename(lvl=level, CT=cell_type) %>%
  arrange(lvl,CT)
CT <- unique(df.cluster$CT)

#=== enrichment analysis:
for(i in 1:length(CT)){
  mydatname <- CT[i]
  tmp <- df.cluster %>% distinct(CT,lvl) %>% filter(CT==mydatname)
  mylvl <- as.character(tmp[1,"lvl"])
  b <- df.cluster %>% 
    filter(CT==mydatname) %>% 
    select(ensgid) %>%
    mutate(mydat=1)
  
  #=== INNER join with test list background, count "R"
  c <- inner_join(background, b, by = "ensgid") 
  R <- as.numeric(sum(c$mydat))
  
  #=== get the overlap, LEFT to preserve full geneset
  # add count vars, fix NAs
  j <- left_join(genesets, c, by = "ensgid") %>%
    select(-ensgid) %>%
    mutate(cnt = 1) %>%
    replace_na(list(mydat=0))
  
  #=== get P-hyper
  l <- j %>%
    group_by(group, subgroup, genesetID) %>% 
    summarise(nGenes=sum(cnt),
              Ngene = Ngene,
              R = R,
              x=sum(mydat)) %>%
    ungroup() %>%
    filter(nGenes>=0)
  
  l$PH <- phyper(l$x-1, l$R, l$Ngene-l$R, l$nGenes, lower.tail=FALSE)
  
  l <- l %>% 
    group_by(group) %>%
    adjust_pvalue(p.col = "PH", output.col = "P.fdr.group", method = "fdr") %>%
    ungroup()
  
  #=== select sig. results
  g <- l %>% filter(P.fdr.group<=GSA.FDRthresh) 
  g$annotation <- mydatname
  g$annotation_lvl <- mylvl

  #=== store results in one data frame (differentiated by "annotation" column)
  if(i==1){
    myres <- g
  }else{
    myres <- rbind(myres, g)
  }
}

#=== get meanings of genesets
genesetsA <- fread("./workflow/genesetsA.tsv.gz")

myres <- myres %>%
  left_join(genesetsA, by="genesetID") %>% 
  filter(subgroup!="HPO")  %>%
  separate(col=geneset, sep=" ", into=c("ID","description"), remove = F)

myres2 <- myres %>%
  left_join(df.color, by=c("annotation"="CellType"))
scores <- setNames(-log10(myres2$P.fdr.group), myres2$ID)

# short CT names
myres2$annotation <- gsub("_neurons","",myres2$annotation)
myres2$annotation <- gsub("_Layer","",myres2$annotation)
myres2$annotation <- gsub("Oligodendrocyte_progenitor_cells","OPC",myres2$annotation)
myres2$annotation <- gsub("Excitatory","Ex",myres2$annotation)
myres2$annotation <- gsub("Inhibitory","Inh",myres2$annotation)

#== loop over the GO categories and make dot plots
for(myGO in c("BP", "CC", "MF")){
  simMatrix <- calculateSimMatrix(myres2$ID,
                                   orgdb="org.Hs.eg.db",
                                   ont=myGO,
                                   method="Rel")
  reducedTerms <- reduceSimMatrix(simMatrix,
                                   scores,
                                   threshold=0.7,
                                   orgdb="org.Hs.eg.db")
  dat2 <- myres2 %>% 
    select(annotation,geneset,P.fdr.group,CT_color,ID,subgroup) %>%
    inner_join(reducedTerms,by=c("ID"="go")) 
  
  #- plot
  tmp <- dat2 %>% distinct(annotation, CT_color) 
  dat2$annotation <- factor(dat2$annotation, levels=tmp$annotation)
  p <- ggplot(data=dat2,
            aes(x=annotation, y=geneset, color=annotation, size=-log10(P.fdr.group))) +
  #geom_point(shape=15) +
  geom_point() +
  theme_bw() +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5)) +
  scale_color_manual(values=tmp$CT_color) +
  facet_grid(rows = vars(parentTerm), scales = "free",space = "free") +
  theme(strip.text.y = element_blank()) +
  xlab("") + ylab("") +
  labs(title=paste("GO:", myGO,sep=""))
  
  plotname <- paste(MYOUTDIR,"figS18_",myGO,"_TWAS",TWAS.FDRthresh,"_GSA",GSA.FDRthresh,"_",lvl,".pdf", sep="")
  if(myGO=="BP"){
    ggsave(plot = p, filename=plotname, width=12, height=16.5)
  }elseif(myGO=="CC"){
    ggsave(plot = p, filename=plotname, width=10, height=10)
  }elseif(myGO=="MF"){
    ggsave(plot = p, filename=plotname, width=11, height=7)
  }
}

```
  
--- script ends ---  